---
title: "Point Processes (Homework 2)"
author: Gregory Matesi
output: 
    pdf_document :
      latex_engine : xelatex
---

```{r}
library(spatstat)
library(splancs)
```
# Problem 1.

The spatstat package includes a data set named japanesepines.  The data give 
the locations of Japanese black pine saplings in a square sampling region in a natural 
forest. The observations were originally collected by Numata (1961). 

## a

**Plot the data.  Comment on what you see and whether this looks compatible 
with CSR.**

```{r}
data("japanesepines")

#class(japanesepines)
#?japanesepines
#?ppp.object

plot(japanesepines)
```

I am seeing a seemingly random scatter of data with natural looking clusters of event locations. Initially, this looks consistent with Complete Spatial Randomness.

## b

**Estimate and create a contour plot of the intensity function (use Scott’s rule 
for the bandwidth in each direction).  Comment on the plot in relation to 
potential departures from CSR.**

Scott's rule: $\hat{b}_u = \hat{\sigma}_u N^{-1/(dim+4)}$ where $\hat{\sigma}_u$
 is the sample standard deviation of the $u$-coordinates, $N$ is the number of events in the study area, and $dim$ is the dimension of the study aread.
 
$$\hat{b}_u = \hat{\sigma}_u N^{-1/(dim+4)},$$
$$\hat{b}_v = \hat{\sigma}_v N^{-1/(dim+4)}$$
```{r}
sigma_u <- sd(japanesepines[[3]])
sigma_v <- sd(japanesepines[[4]])
dim_uv <- 2
N <- japanesepines[[2]]

bandwidth_u <- sigma_u * N**( -1 / (dim_uv + 4) )
bandwidth_v <- sigma_v * N**( -1 / (dim_uv + 4) )

bandwidth_u_2 <- sd(japanesepines$x) * japanesepines$n ** ( -1 / (dim_uv + 4) )
bandwidth_v_2 <- sd(japanesepines$y) * japanesepines$n ** ( -1 / (dim_uv + 4) )

bandwidth_u; bandwidth_v
bandwidth_u_2 == bandwidth_u; bandwidth_v_2 == bandwidth_v_2
rm(bandwidth_u_2, bandwidth_v_2)
```

```{r}
intensity.function <- density(japanesepines, sigma = c(bandwidth_u, bandwidth_v))
intensity.function
```

```{r}
par(mfrow = c(1, 2))
persp(intensity.function, theta = 45, phi = 35, xlab = "u", ylab = "v", zlab = "density", main = "Estimated intensity function")
contour(intensity.function, xlab = "u", ylab = "v", main = "Japanese Pines", nlevels = 15)
points(japanesepines, pch = ".")
```

The estimated intensity function seems to vary over space. This suggest we have more pines growing at the modes of the intensity function and less pines growing between the modes. This could indicate that we may have a departure from Complete Spatial Randomness in these data. Further analysis would be needed to confirm this.

## c

**Construct an $\hat{L}(h)-h$ plot for the data with 95% acceptance/non-rejection 
envelopes.  You do not need to specify *r* (the vector of distances) in the 
function—simply use the default values.  What are your conclusions?**

```{r}
lplot <- function(x, nsim = 500, level = 0.95,
                  correction = "Ripley", ...) {
  lobs <- spatstat.core::Lest(x, correction = correction, ...)

  # lambda <- summary(x)$intensity
  win <- x$window
  lsim <- pbapply::pblapply(1:nsim, FUN = function(i) {
    # xsim <- spatstat::rpoispp(lambda, win = win)
    # generate n event locations over study area under CSR
    xsim <- spatstat.core::runifpoint(n = x$n, win = win)
    # estimate L for simulated point pattern
    spatstat.core::Lest(xsim, correction = correction, ...)
  })

  r <- lobs$r # get distances
  obs <- lobs$iso # get estimated l for observed
  # get estimated l for each simulated data set
  sim <- sapply(lsim, getElement, "iso")
  # apply the min function to each row  (MARGIN = 1) of sim
  # gets pointwise minimum for simulated data
  # at each distance.  do same for max, quantiles, median
  lo <- apply(sim, MARGIN = 1, FUN = min, na.rm = TRUE)
  hi <- apply(sim, MARGIN = 1, FUN = max, na.rm = TRUE)
  alpha <- 1 - level
  qlo <- apply(sim, MARGIN = 1, FUN = quantile,
               prob = alpha/2, na.rm = TRUE)
  qhi <- apply(sim, MARGIN = 1, FUN = quantile,
               prob = 1 - alpha/2, na.rm = TRUE)
  med <- apply(sim, MARGIN = 1, FUN = median, na.rm = TRUE)
  # construct empty plot of the right size
  plot(range(r), c(min(c(lo, obs) - r, na.rm = TRUE),
                   max(c(hi, obs) - r, na.rm = TRUE)),
       type = "n",
       xlab = "distance", ylab = "L(distance) - distance")
  # plot different statistics with different styles/thickness
  lines(r, obs - r, lwd = 2)
  #lines(r, lo - r, lty = 2)
  #lines(r, hi - r, lty = 2)
  lines(r, qlo - r, lty = 1)
  lines(r, qhi - r, lty = 1)
}
```

```{r}
set.seed(1)
lplot(japanesepines, level = .95)
```
Based on this plot, the %95 percent quantile plot provides no evidence of departure from Complete Spatial Randomness across any spatial scale from 0 to 0.25. The lines fall within our envelope across the entire spectrum of spatial scales tested.

## d

**Perform a global test to determine whether there is a departure from CSR in 
the data set at a significance level of 0.05.   In this case, the maximum spatial 
scale to consider (h\*) should be 0.25.  What is your conclusion?**

```{r}
# test hypothesis that affected event locations are
# clustered at any scale for 0 <= h <= 2000
# in rectangular domain
numsim <- 499
r <- seq(0, .25, len = 201)
Tobs <- max(spatstat.core::Lest(japanesepines, r = r, correction = "Ripley")$iso - r)
# relabel affected events,
# then compute max(Lhat(h) - h)
# for relabeled data
Tsim <- pbapply::pbsapply(1:numsim, FUN = function(i) {
  max(spatstat.core::Lest(japanesepines[sample.int(japanesepines$n, size = japanesepines$n), 2:3], r = r,
           correction = "Ripley")$iso - r)
})

length(japanesepines[sample.int(japanesepines$n, size = japanesepines$n), 2:3])
japanesepines[sample.int(japanesepines$n, size = japanesepines$n), 2:3][2]
# proportion of simulated test statistics
# as extreme as one observed
# the observed pattern is relative consistent with a random labeling of affected
pval1 <- mean(c(Tsim, Tobs) >= Tobs)
pval1
```

This p-value, `r pval1`, of well over a general rejection threshhold of 0.05. using this test, we do not hace sufficient evidence to conclude that there is a global departure from complete spacial randomness.
We could also say, based on this test, on a global scale, we do not see more clustering in the observed data than we would expect to see under complete spacial randomness.

## e

**Make conclusions about whether there is evidence of departure from CSR, 
where the departures may occur, and at what scale they may occur based on 
your information from the previous three problem parts.**

Based on the quantile L plot and the global test for departure from CSR, we do not have sufficient evidence to conclude that the Japanese pines are any more clustered than they would be if they were distributed across this study area under Complete Spacial Randomness.


# Problem 2

Repeat the analysis from problem 1 for the redwood data in the spatstat package. 
The data represent the locations of 62 seedlings and saplings of California redwood 
trees in a square sampling region. They originate from Strauss (1975); the present 
data are a subset extracted by Ripley (1977) in a subregion that has been rescaled to 
a unit square.  The coordinates are rounded to the nearest 0.01 units, except for one 
point which has an x coordinate of 0.999, presumably to ensure that it is properly 
inside the window. 

## a

```{r}
data("redwood")
plot(redwood)
```

In this plot, it seems like we have a regular spread of clusters of event locations. If we zoomed in on each of these 7 or so clusters, however, I think that the event locations within each of these clusters would look more or less regular.

## b

```{r}
sigma_u <- sd(redwood[[3]])
sigma_v <- sd(redwood[[4]])
dim_uv <- 2
N <- redwood[[2]]

bandwidth_u <- sigma_u * N**( -1 / (dim_uv + 4) )
bandwidth_v <- sigma_v * N**( -1 / (dim_uv + 4) )

bandwidth_u_2 <- sd(redwood$x) * redwood$n ** ( -1 / (dim_uv + 4) )
bandwidth_v_2 <- sd(redwood$y) * redwood$n ** ( -1 / (dim_uv + 4) )

bandwidth_u; bandwidth_v
bandwidth_u_2 == bandwidth_u; bandwidth_v_2 == bandwidth_v_2
rm(bandwidth_u_2, bandwidth_v_2)
```


```{r}
intensity.function <- density(redwood, sigma = c(bandwidth_u, bandwidth_v))
intensity.function
```

```{r}
par(mfrow = c(1, 2))
persp(intensity.function, theta = 45, phi = 35, xlab = "u", ylab = "v", zlab = "density", main = "Estimated intensity function")
contour(intensity.function, xlab = "u", ylab = "v", main = "Redwood", nlevels = 15)
points(redwood, pch = ".")
```

The conclusion for tis plot is similar to the conclusion for part b of problem 1. In my opinion, however, the variation of the intesity function across the study area in this problem looks more severe. Here, we are intensity function values ranging from 10 to 110. In problem 1, we were seeing intensity function values ranging from 40 to 105 or so.

## c

```{r}
set.seed(1)
lplot(redwood, level = .95)
```
This plot indicates that we may be seeing more clustering than under the null hypothesis of CSR in spatial scales from 0.03 to 0.2.
We have evidence of clustering relative to CSR in the spatial range (0.03, 0.3).

## d

```{r}
# test hypothesis that affected event locations are
# clustered at any scale for 0 <= h <= 2000
# in rectangular domain
numsim <- 499
r <- seq(0, .25, len = 201)
Tobs <- max(spatstat.core::Lest(redwood, r = r, correction = "Ripley")$iso - r)
# relabel affected events,
# then compute max(Lhat(h) - h)
# for relabeled data
Tsim <- pbapply::pbsapply(1:numsim, FUN = function(i) {
  max(spatstat.core::Lest(redwood[sample.int(redwood$n, size = redwood$n), 2:3], r = r,
           correction = "Ripley")$iso - r)
})

# proportion of simulated test statistics
# as extreme as one observed
# the observed pattern is relative consistent with a random labeling of affected
pval2 <- mean(c(Tsim, Tobs) >= Tobs)
pval2
```
Based on the global test of clustering relative to CSR above, we obtain a p-value of `r pval2`. This does not indicate that we are seeing more clustering relative to CSR on a global scale (across the entire study area).

## e

Based on the above analyses, we can determine that there is a departure from Complete Spatial Randomness on spatial scales ranging from 0.03 to 0.2. There is not, however, evidence for departure from CSR on a global scale.

# Problem 3

The splancs package is another popular (though older) package for statistical 
analysis of spatial point patterns in R.  Reproduce the top panel in Fig 5.13 using the 
splancs package.  The khat and csr functions should come in handy. 

```{r}
library(splancs)
```

```{r eval=FALSE}
?khat
?csr
```

```{r}

```
