---
title: "hw3"
output: 
    pdf_document :
      latex_engine : xelatex
---

# Problem 1

The *Urkiola* data set in the **spatstat** package contains locations of birch (*Betula celtiberica*) and oak (*Quercus robur*) trees in a secondary wood in Urkiola NAtural Park (Basque country, Northern Spain). They are part of a more extensive dataset collected and analysed by Laskurain (2008). The coordinates of the trees are given in meters. Let the "oak" trees be the cases and birch trees be the controls.



```{r}
my.path <- "~/../Desktop/SpatialStatisticsClass/"
set.seed(1)
library(spatstat)
library(smacpod)

data("urkiola")
str(urkiola)
class.urkiola <- class(urkiola)

rm(class.urkiola)
oak <- which(urkiola$marks == "oak")
#birch <- which(urkiola$marks == "birch")
```

## part a

**Create a plot of the point pattern that distinguishes between the two types of trees. Do you notice any evidence of clusters? Explain.**

```{r}
plot(urkiola)
```



## Part b

**Create contout plots for the spatial density function for the oak and birch trees. Use bandwidth of 14.5 for oak and 15 for birch. Do any differences jump out at you?**

```{r}
oakbw <- bw.scott(urkiola[oak,])
birchbwd <- bw.scott(urkiola[-oak,])

oakbw
birchbwd

# estimate densities of cases and controls, respectively,
# using bandwidth of 700
f.density = spdensity(urkiola[oak, ], sigma = 14.5)
g.density = spdensity(urkiola[-oak,], sigma = 15)

# contour plot of f700 with title
contour(f.density, nlevels = 15, main = "")
title("Oak, bandwidth = 700")
# contour plot of g700, with title
contour(g.density, nlevels = 15, main = "")
title("Birch, bandwidth = 700")
```

## Part c

**Estimate the log ratio of the spatial density function of the oak relative to the birch trees (r(s)) using a bandwidth of 16. Create a contout plot of the log ratio. Which areas are most consistent with the belief that the spatial densities of the two trees are the same?**

```{r eval}
# log ratio of spatial densities
r16 = logrr(urkiola, case = 2, sigma = 16)
r16


#contour plot of r700 (lty and lwd determined
# by experimentation)
contour(r16, lty = c(1, 1, 1, 1, 2, 1),
        lwd = c(1, 1, 1, 1, 1, 2), main = "")
title("Gaussian kernel, Bandwidth = 16")
```

## Part d

**construct pointwise 95% non-rejection limits of r(s) using 499 datasets simulated under the random labelling hypothesis. Plot the regions above and below the limits in different colors. Overlay the contour plot of $\tilde{r}(s)$. What can you conclude?

```{r}
# construct 95% tolerance envelopes for log relative risk
# when bandwidth = 350

# 
# urkiola_renv = logrr(urkiola, case = 2, nsim = 999, sigma = c(14.5, 15),
#                 level = 0.95)
# save(urkiola_renv, file = "~/../Desktop/SpatialStatisticsClass/chapter6_casecontrol/urkiola_renv.rda")

load("~/../Desktop/SpatialStatisticsClass/chapter6_casecontrol/urkiola_renv.rda")
# image plot showing regions where r350 is outside
# tolerance envelopes
plot(urkiola_renv)

# a better color scale (in my opinion)
# making it easier to distguish the clusters of cases relative
# to controls (red) and vice versa (blue)
grad = gradient.color.scale(min(urkiola_renv$v, na.rm = TRUE), max(urkiola_renv$v, na.rm = TRUE))
plot(urkiola_renv, col = grad$col, breaks = grad$breaks)


```

## Part e

**Perform a global test of clustering using $\tilde{r}(s)$. Is there convincing evidence of clustering of one group relative to the other for at least one location in the study area?**

```{r}
logrr.test(urkiola_renv)
```

Yes!

## Part f

**Construct a plot for the difference in K functions for oak and birch trees. Also include min/max envelopes using 499 simulated datasets under the random labelling hypothesis. Also construct 95% pointwise non-rejection envelopes. Also include the mean difference from the simulations. Label all these things in a legend.** \n
**Are there any spatial scales in which the oak trees are more clustered in relation to the birch trees to what we expect under the random labelling hypothesis (or vice versa)? If so, what scales?**

```{r eval=FALSE}
my.nsim <- 499
kd_urkiola <- kdest(urkiola, case = "oak")
plot(kd_urkiola, main = "Just KD_+ estimate")
plot(kd_urkiola, cbind(iso, theo) ~ r, legend = FALSE, main = "")

# kdenv_urkiola = kdest(urkiola, case = "oak", nsim = my.nsim,
#               #r = seq(0, 100, len = 201),
#               level = 0.95)
# save(kdenv_urkiola, file = "~/../Desktop/SpatialStatisticsClass//chapter6_casecontrol/kdenv_urkiola.rda")

load("~/../Desktop/SpatialStatisticsClass//chapter6_casecontrol/kdenv_urkiola.rda")

plot(kdenv_urkiola, ylab = "difference in K functions",
     xlim = c(0, 40)
     ); legend("topleft", legend = c("obs", "avg", "max/min env", "95% env"),
       lty = c(1, 2, 1, 2), col = c("black", "red", "gray56", "lightgrey"),
       lwd = c(1, 1, 10, 10)); abline(v=14)
```

## Part g

**Perform a global test of clustering for oak trees relative to birch trees using the $KD_+$ statistic. Interpret the results.**

```{r}
kdplus.test(kdenv_urkiola)

```

Yes!



